Variational oblique predictive clustering trees

### 2025-06-19 reports/paper/response_to_reviewers.tex

\documentclass [a4paper,11pt]{article}
\usepackage{enumitem}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
%\usepackage{cite} 
%\usepackage[sort&compress,square,numbers]{natbib}
\usepackage[backend=biber,sorting=none]{biblatex}
\addbibresource{references.bib}
\usepackage{color}
\setlength{\itemsep}{0pt plus4pt minus2pt}
\renewcommand{\labelitemi}{\Alph{enumi}}
\setlength{\headheight}{0mm} \setlength{\topskip}{0mm}
\setlength{\textheight}{24cm} \setlength{\topmargin}{-15mm}
\setlength{\textwidth}{16cm}\setlength{\oddsidemargin}{0cm}\setlength{\evensidemargin}{0cm}
\setlength{\parindent}{0mm} \setlength{\parskip}{2mm}
\usepackage{url}
\usepackage{amsthm}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{eurosym}
\usepackage{enumitem}
\usepackage{graphicx}
\usepackage[colorinlistoftodos]{todonotes}
\usepackage{booktabs}
\usepackage[version=4]{mhchem}
\usepackage[subrefformat=parens,labelformat=parens]{subfig}
\usepackage{comment}
\usepackage{bm}
\graphicspath{ {./}{./figs/}{./figs/laprint_figs/} }

\definecolor{SteelBlue}  {rgb}{0.275,0.51,0.706}

\usepackage{xr}
\externaldocument[m-]{main}

\newtheoremstyle{blue}% name
{10pt}%      Space above
{3pt}%      Space below
{\color{SteelBlue}}%         Body font
{}%         Indent amount (empty = no indent, \parindent = para indent)
{\bfseries}% Thm head font
{:}%        Punctuation after thm head
{.5em}%     Space after thm head: " " = normal interword space;
%       \newline = linebreak
{}%         Thm head spec (can be left empty, meaning `normal')

\theoremstyle{blue}
\newtheorem{question}{Comment}[section]

\definecolor{green-red}  {rgb}{0,0.,1}
\newtheoremstyle{blue}% name
{10pt}%      Space above
{3pt}%      Space below
{\color{green-red}}%         Body font
{}%         Indent amount (empty = no indent, \parindent = para indent)
{\bfseries}% Thm head font
{:}%        Punctuation after thm head
{.5em}%     Space after thm head: " " = normal interword space;
%       \newline = linebreak
{}%         Thm head spec (can be left empty, meaning `normal')

\theoremstyle{blue}
\newtheorem{general}{General Comment}[section]

\newtheoremstyle{note}% name
{3pt}%      Space above
{3pt}%      Space below
{}%         Body font
{}%         Indent amount (empty = no indent, \parindent = para indent)
{\bfseries}% Thm head font
{:}%        Punctuation after thm head
{.5em}%     Space after thm head: " " = normal interword space;
%       \newline = linebreak
{}%         Thm head spec (can be left empty, meaning `normal')

\theoremstyle{note}
\newtheorem{answer}{Response}[section]

\newtheoremstyle{note}% name
{3pt}%      Space above
{10pt}%      Space below
{}%         Body font
{}%         Indent amount (empty = no indent, \parindent = para indent)
{\bfseries}% Thm head font
{:}%        Punctuation after thm head
{.5em}%     Space after thm head: " " = normal interword space;
%       \newline = linebreak
{}%         Thm head spec (can be left empty, meaning `normal')

\theoremstyle{note}
\newtheorem{general_answer}{General Response}[section]

\newcommand{\todopb}[2][]
{\todo[backgroundcolor=blue!25, #1]{#2}}

\usepackage[normalem]{ulem}
\newcommand{\mkom}[1]{{\color{purple} #1}}
\newcommand{\mnew}[1]{{\color{Tan} #1}}
\newcommand{\mpop}[2]{{\color{orange}\sout{#1}\color{purple}{#2}}}
\newcommand{\mrev}[2]{{\color{red}\footnote{#1} #2 }}

\usepackage[nomargin,inline,marginclue,draft]{fixme}
 \fxsetup{theme=color, mode=multiuser} 
 \FXRegisterAuthor{pb}{apb}{\color{red}PB} 
 \FXRegisterAuthor{mp}{amp}{\color{red}mp}

\usepackage{xr}
\externaldocument[M-]{detcoupling}

\begin{document}

	\textbf{Detailed Response to Reviewers}\\[10pt]

%	{\large \textbf{Reply to the reviewers}}
%	This document contains the responses to the reviewers' questions.
%	In addition to the revised manuscript, we have added a marker revised manuscript that includes color-highlighted changes marked with a reference to the reviewers' questions.

%	The analysis of the mutual dependence of $(N,L)$ and the dependence on the parameters $k_1$ and $k_2$ are presented as Supplementary material.
%	We have also moved the original Appendix~A as a part of the Supplementary material.

	\section{Reviewer \#1:}

%{\color{blue}Section I)}

\begin{question}
Correct small typographical issues (e.g., replace "it's" with "its" in formal writing).
\end{question}

\begin{answer}

Thank you. This is now corrected.

\end{answer}

%{\color{blue}Section II)}

\begin{question}

Improve Figure 1's caption to clearly distinguish ensemble vs. variational trees.

\end{question}

\begin{answer}

The caption has been improved as per your suggestion.

\end{answer}

\begin{question}

Consider summarizing model characteristics in a comparative table in the Related Work section.

\end{question}

\begin{answer}

We have now included a comparative table in the Related Work section clearly summarising and comparing the key characteristics of the VSPYCT method and the other predictive clustering methods in the literature.

\end{answer}

\begin{question}

Clarify "structured output prediction" early for readers not familiar with the term.

\end{question}

\begin{answer}

We clarified the term Structured Output Prediction (SOP) explicitly in the Introduction section by providing a concise definition for readers who may not be familiar with this concept.

\end{answer}

\begin{question}

Include explanation of $\Omega_{obs} = \Omega (X,Y,w,b)/2$ mentioned in Algorithm1 and why this choice is made.

\end{question}

\begin{answer}

We have clarified this in the manuscript (inside Algorithm 1), explicitly stating that the factor of 1/2 in $\Omega_{obs} = \Omega (X,Y,w,b)/2$ is introduced to ensure that it is consistently lower than the current impurity $\Omega (X,Y,w,b)$. This guarantees that the optimisation is biased toward impurity reduction at each step, thereby encouraging effective split selection. The choice helps to drive the learning dynamics in the right direction by maintaining a useful margin between the observed and expected impurities.

\end{answer}

\begin{question}

Although the paper acknowledges the added complexity of variational inference compared to deterministic tree splits, it lacks empirical evidence regarding training time
or scalability.

\end{question}

\begin{answer}

Thank you for highlighting this important aspect. While we have not provided empirical training-time measurements due to their high dependence on specific implementations and hardware, we have explicitly derived and clearly presented the theoretical computational complexity of our proposed VSPYCT model compared to the deterministic SPYCT. The added complexity due to the variational inference framework is detailed explicitly in the manuscript. To clarify the limitations, we have now clearly stated in the manuscript that systematic empirical evaluations of training time and scalability were outside of the scope of this study, due to the above-mentioned reasons.

\end{answer}

\begin{question}

The model captures parameter uncertainty well, but structural uncertainty (e.g., tree topology) is briefly mentioned as future work.

\end{question}

\begin{answer}

Thank you for this important suggestion. The Discussion section now highlights structural uncertainty as a significant area for future research. Specifically, we mentioned potential strategies such as incorporating structural priors or employing ensemble-like approaches within the variational framework.

\end{answer}

\begin{question}

There is no mention of code availability, parameter tuning strategies, implementation frameworks or datasets . Consider including a statement on code or data
availability to enhance reproducibility.

\end{question}

\begin{answer}

We have now included a clear statement in the manuscript with a link to our code implementation.

\end{answer}

\begin{question}

While feature importance and visualisations are well described, a brief practical commentary on how domain experts might interact with or interpret the VSPYCT outputs
would be valuable.

\end{question}

\begin{answer}

We have explicitly added a brief practical commentary in the Interpretability subsection, clearly outlining how domain experts (e.g., clinicians, employment analysts, or material scientists) might effectively interact with and interpret VSPYCT outputs for informed decision-making in their respective fields.

\end{answer}

	\section{Reviewer \#2:}

\begin{question}

The novelty of the work should be clearly emphasized in the last paragraph of the introduction.

\end{question}

\begin{answer}

Thank you for this important remark. We have now explicitly emphasised the novelty of our approach in the final paragraph of the Introduction, clearly highlighting the integration of Bayesian inference with oblique splits, the direct quantification of uncertainty, and the unique capability to maintain interpretability while achieving competitive predictive performance.

\end{answer}

\begin{question}

Too many parameters have been examined in the study, but few results are presented in the Conclusion section. Please add a few points in the Conclusion section.

\end{question}

\begin{answer}

We have enhanced the Conclusion section with specific results summarizing the predictive performance of VSPYCT compared to benchmark methods. Additionally, we explicitly discussed key findings, limitations, and recommendations for future research clearly.

\end{answer}

\begin{question}

All parameters and symbols mentioned in this manuscript must be defined within the manuscript or in a separate list of nomenclature.

\end{question}

\begin{answer}

Thank you for this. All parameters and symbols used throughout the manuscript have now been explicitly defined at the point of introduction within the main text. We have ensured that no undefined notation remains, making the manuscript self-contained without the need for a separate nomenclature table.

\end{answer}

\begin{question}

The introduction is not well organized. The research gap and novelty of the research are missing. The author should clearly explain what has done previously and what is the research gap. The authors did not explain the cited articles properly.

\end{question}

\begin{answer}

Thank you for highlighting this issue. We reorganized the Introduction section by explicitly identifying the research gap—particularly emphasizing the lack of integrated uncertainty quantification and interpretability in previous studies. We have also clarified the context and relevance of the cited literature.

\end{answer}

\begin{question}

For the validity of the results, the study needs to be validated with previous studies graphically.

\end{question}

\begin{answer}

Thank you for highlighting this important point.
To clarify, our experimental design already includes comparisons with state-of-the-art models from previous literature, such as standard SPYCT, ensembles of SPYCTs, OPCT, and ensembles of PCTs.
These baseline models have been extensively evaluated and validated in previous published studies (Stepišnik and Kocev, 2021; Kocev et al., 2013; Stepisnik et al., 2020), and we explicitly included them in our empirical analyses. The ranking based on the predictive performance is given in Figure 2.

\end{answer}

\begin{question}

What is the novelty of this research?

\end{question}

\begin{answer}

The novelty of our research lies in the unique integration of variational Bayesian inference with oblique predictive clustering trees. This allows us to directly quantify uncertainty within a single interpretable tree model, achieving predictive performance comparable or superior to state-of-the-art ensemble methods, which has not been explored in previous studies.

\end{answer}

\begin{question}

The result and discussion section should be improved from the physical point of view.

\end{question}

\begin{answer}

We have introduced a new subsection "Practical Interpretation of Results" in the Results and Discussion section. This clearly addresses the physical and practical significance of our findings, explicitly linking the predictive outcomes to real-world contexts such as material properties, medical diagnoses, and employment data.

\end{answer}

\begin{question}

How were similarity parameters calculated?

\end{question}

\begin{answer}

We have explicitly clarified in the Methodology section how similarity parameters, specifically feature importance, are calculated. These parameters are computed based on the ratio of the expected values of the weights to their variances, which integrates both the magnitude and uncertainty of each feature's contribution to the predictions.

\end{answer}

\begin{question}

Include justifications for the choice of specific parameters and models.

\end{question}

\begin{answer}

We have clearly included justifications for our chosen parameters and models within the Methodology section. Specifically, we explained our choice of Gaussian priors for computational tractability, the selection of the ELBO optimisation approach, and parameters used in Monte Carlo sampling based on trade-offs between predictive accuracy and computational efficiency.

\end{answer}

\begin{question}

It is recommended to noticeably highlight the advantages, disadvantages, and limitations of the proposed approaches.

\end{question}

\begin{answer}

We have clearly added an explicit and structured discussion of the advantages, disadvantages, and limitations of the VSPYCT model within the Discussion subsection. Advantages include interpretability and uncertainty quantification, while disadvantages primarily involve computational complexity and potential scalability concerns.

\end{answer}

\begin{question}

It is necessary to incorporate further analysis and discussion comparing the current findings with previous publications. There must be an explanation for any differences between the present results and the data that has been published.

\end{question}

\begin{answer}

\end{answer}

\begin{question}

It is recommended to add a table representing all the investigated cases in the manuscript; please show the values/range of each tested parameter.

\end{question}

\begin{answer}

We have now added a comprehensive table under the "Experimental Setting" section, summarising all parameters investigated, their ranges, and their specific relevance. Furthermore, we explicitly stated the final selected parameter values used throughout the experiments clearly below the table.

\end{answer}

\begin{question}

The conclusion section should be better organized and it is recommended to include quantitative results. More specific aspects regarding the limitations, recommendations, and future works can be stated in the conclusion section. It is recommended to include qualitative and quantitative results.

\end{question}

\begin{answer}

We have reorganised and enhanced our Conclusion section to explicitly include both qualitative and quantitative summaries of our key findings.

\end{answer}

\begin{question}

Ensure that all sources and data are properly cited throughout the manuscript. For instance, the following papers could be helpful in improving the literature

\end{question}

\begin{answer}

Thank you for this suggestion. We have now cited the recommended references within our manuscript.

\end{answer}

\end{document}

### 2025-06-19 reports/paper/markup_vspyct.tex
First revision

\Glspl{spyct} are semi-supervised multi-target prediction models mainly used for \gls{sop} problems.
They are computationally efficient and when combined in ensembles they achieve state-of-the-art results.
However, one major issue is that it's challenging to interpret an ensemble of \glspl{spyct} without the use of a model-agnostic method.
We propose variational oblique predictive clustering trees, which address this challenge.
The parameters of each split node are treated as random variables, described with a probability distribution, and they are learned through the Variational Bayes method.
We evaluate the model on several benchmark datasets of different sizes.
The experimental analyses show that a single \gls{vspyct} achieves competitive, and sometimes better predictive performance than the ensemble of standard \glspl{spyct}.
We also present a method for extracting feature importance scores from the model.
Finally, we present a method to visually interpret the model's decision making process through analysis of the relative feature importance in each split node.
\end{abstract}
%\maketitle
\end{frontmatter}
\glsresetall

@@

In the evolving field of machine learning, the fusion of semi-supervised learning and decision tree models has led to the development of models like \glspl{pct}~\citep{Kocev_2013} and their advanced variant, \glspl{spyct}~\citep{Stepi_nik_2021}.
These models leverage labeled and unlabelled data to enhance predictive accuracy and efficiency.
\Glspl{spyct}, in particular, have pioneered the use of oblique splits for \gls{sop}, allowing for more intricate decision boundaries beyond the capabilities of traditional axis-parallel splits.
Despite their advantages, a significant challenge with \glspl{spyct} is their reliance on ensemble methods for optimal predictive performance.
Ensemble methods, such as random forests and gradient-boosted trees, achieve high predictive performance by reducing variance and/or bias through the aggregation of diverse models~\cite{Breiman_2001, Kocev_2020, Friedman_2001}.
These methods ensure robustness by capturing slightly different patterns or residuals from the data.
While effective in improving accuracy, ensembles tend to obscure the model's interpretability—a fundamental attribute of decision trees—and do not inherently provide a means to quantify prediction uncertainty, which is crucial for informed decision-making across various applications.

@@

\begin{figure}[h!]
    \centering
    \includegraphics[width=1.0\textwidth]{main_flowchart.pdf}
    \caption{Ensemble of trees vs. Variational tree}
    \label{fig:main_flowchart}
\end{figure}

Our proposed \gls{vspyct} framework diverges from traditional ensemble approaches by directly incorporating Bayesian principles into oblique decision trees' architecture.
This model avoids conventional ensemble strategies, which aggregate multiple trees' outputs, for a probabilistic treatment of model parameters, offering refined mechanisms for uncertainty quantification and predictive performance.
The core of \gls{vspyct}'s innovation lies in applying the variational Bayes method for optimising parameters defining the oblique splits, and modelling these parameters as random variables with probability distributions.
This represents a paradigm shift toward a probabilistic understanding of decision-making processes within tree-based models, significantly enhancing interpretability, especially in uncertainty estimation domains.
It sets a new benchmark for interpretable, efficient, and reliable machine learning models.
\pbnote{Check Bayesian Additive Regression Trees (BART). Maybe add a comment/reference what is different}

@@

@@

%references for VB applications in ML and Survival analysis models. Cite pavles paper.

@@

@@

@@

    \Procedure{LearnSplit}{$\mathcal{D}, \theta, E, \lambda, \beta, \sigma$}
        \State Initialise $\Theta = \{\boldsymbol{\mu}_w, \boldsymbol{\Sigma}_w, \mu_b, \sigma_b^2\}$ \Comment{Initialise variational parameters}
        \State Initialise optimiser: $\alpha \gets \lambda$
        \For{$e \in \{1, \dots, E\}$} \Comment{Iterate over epochs}
            \For{each mini-batch $\mathcal{B} \subseteq \mathcal{D}$ of size $\beta$}
                \State Sample $\mathbf{w} \sim \mathcal{N}(\boldsymbol{\mu}_w, \boldsymbol{\Sigma}_w)$, $b \sim \mathcal{N}(\mu_b, \sigma_b^2)$
                \State Compute impurity $\Omega(\mathbf{X}, \mathbf{Y}; \mathbf{w}, b)$ as:
                \[
                \Omega(\mathbf{X}, \mathbf{Y}; \mathbf{w}, b) = \sum_{\text{right}} \rho(\mathbf{w}^\top \mathbf{x} + b) \cdot \text{Var}_\rho(\mathbf{Y}) + \sum_{\text{left}} (1 - \rho(\mathbf{w}^\top \mathbf{x} + b)) \cdot \text{Var}_\lambda(\mathbf{Y})
                \]
                \State Set the observed impurity value: $\Omega_\text{obs} = \frac{\Omega(\mathbf{X}, \mathbf{Y}; \mathbf{w}, b)}{2}$
                \State Condition on the observed impurity using:
                \[
                \log p(\Omega_\text{obs}|\Omega(\mathbf{X}, \mathbf{Y}; \mathbf{w}, b))
                \]
                \State Compute ELBO $\mathcal{L}(\mathcal{B}; \Theta)$ as per~\eqref{eq:elbo}
                \State Compute gradient $\nabla_{\Theta} \text{ELBO}$
                \State Update $\Theta \gets \Theta + \alpha \nabla_{\Theta} \text{ELBO}$ \Comment{Gradient descent update}
            \EndFor
            \If{early stopping criterion satisfied}
                \State \textbf{break}
            \EndIf
        \EndFor
        \State \textbf{return} $\Theta$
    \EndProcedure
\end{algorithmic}
\end{algorithm}

@@

@@

@@

@@

@@

We introduced the \gls{vspyct} model, an approach that combines the benefits of oblique splits, variational Bayes, and Bayesian inference within a single, interpretable predictive framework.
The \gls{vspyct} model addresses the trade-off between the performance of ensemble models and the interpretability of single decision trees, while also tackling the absence of inherent uncertainty quantification in predictions.

@@

The model opens new possibilities for research in semi-supervised learning and Bayesian inference, promising to bridge the gap between theoretical advancements and practical applications.
Future work will focus on further refining the model's performance, exploring the inclusion of soft splits to enable smoother decision boundaries and probabilistic routing, and extending its applications in diverse predictive tasks.

\begin{comment}
\section*{Acknowledgement}
The authors acknowledge the research core funding No.\ P2-0001,  P2-0103 and P2-0383 financially supported by the Slovenian Research Agency.
The authors also acknowledge the funding received from the European Union's Horizon 2020 research and innovation programme project HECAT under grant agreement No.\ 870702.
\end{comment}

>>>>

\Glspl{spyct} are semi-supervised multi-target prediction models mainly used for \gls{sop} problems.
They are computationally efficient and when combined in ensembles they achieve state-of-the-art results.
However, one major issue is that it is challenging to interpret an ensemble of \glspl{spyct} without the use of a model-agnostic method.
We propose variational oblique predictive clustering trees, which address this challenge.
The parameters of each split node are treated as random variables, described with a probability distribution, and they are learned through the Variational Bayes method.
We evaluate the model on several benchmark datasets of different sizes.
The experimental analyses show that a single \gls{vspyct} achieves competitive, and sometimes better predictive performance than the ensemble of standard \glspl{spyct}.
We also present a method for extracting feature importance scores from the model.
Finally, we present a method to visually interpret the model's decision making process through analysis of the relative feature importance in each split node.
\end{abstract}
%\maketitle
\end{frontmatter}
\glsresetall

@@

In the evolving field of machine learning, the fusion of semi-supervised learning and decision tree models has led to the development of models like \glspl{pct}~\citep{Kocev_2013} and their advanced variant, \glspl{spyct}~\citep{Stepi_nik_2021}.
These models leverage labeled and unlabelled data to enhance predictive accuracy and efficiency.
\Glspl{spyct}, in particular, have pioneered the use of oblique splits for \gls{sop}, allowing for more intricate decision boundaries beyond the capabilities of traditional axis-parallel splits.
\added[id=r1, comment=C4]{In this work we use the term SOP to denote predictive tasks where each example’s label is itself a vector, set or hierarchy (e.g., multi-target regression, hierarchical multi-label, multi-output classification).}
Despite their advantages, a significant challenge with \glspl{spyct} is their reliance on ensemble methods for optimal predictive performance.
Ensemble methods, such as random forests and gradient-boosted trees, achieve high predictive performance by reducing variance and/or bias through the aggregation of diverse models~\cite{Breiman_2001, Kocev_2020, Friedman_2001}.
These methods ensure robustness by capturing slightly different patterns or residuals from the data.
While effective in improving accuracy, ensembles tend to obscure the model's interpretability—a fundamental attribute of decision trees—and do not inherently provide a means to quantify prediction uncertainty, which is crucial for informed decision-making across various applications.

@@

\begin{figure}[h!]
    \centering
    \includegraphics[width=1.0\textwidth]{main_flowchart.pdf}
    \caption{Visual contrast between a bagged ensemble of oblique predictive clustering trees (left) and the proposed single Variational SPYCT (right). The shading of hyper-planes illustrates parameter uncertainty that is captured by the variational posterior, rather than by averaging many independent trees.}
    \label{fig:main_flowchart}
\end{figure}

Our proposed \gls{vspyct} framework diverges from traditional ensemble approaches by directly incorporating Bayesian principles into oblique decision trees' architecture.
This model avoids conventional ensemble strategies, which aggregate multiple trees' outputs, for a probabilistic treatment of model parameters, offering refined mechanisms for uncertainty quantification and predictive performance.

\added[id=r2, comment=C4]{Despite the progress in predictive clustering methods, existing approaches either sacrifice interpretability by relying on ensembles or fail to directly incorporate uncertainty quantification.
Recent works, such as \glspl{spyct}~\citep{Stepi_nik_2021} and \glspl{opct}~\citep{Stepisnik_2020}, have addressed some limitations but lack the integrated capability of uncertainty quantification, critical for real-world decision-making contexts.
Hence, there is a clear research gap in developing methods that simultaneously ensure interpretability, accuracy, and robust uncertainty estimation.}

\begin{comment}
The core of \gls{vspyct}'s innovation lies in applying the variational Bayes method for optimising parameters defining the oblique splits, and modelling these parameters as random variables with probability distributions.
This represents a paradigm shift toward a probabilistic understanding of decision-making processes within tree-based models, significantly enhancing interpretability, especially in uncertainty estimation domains.
It sets a new benchmark for interpretable, efficient, and reliable machine learning models.
\end{comment}

\added[id=r2, comment=C1C6]{In contrast to prior work, \glspl{vspyct} combine oblique splits from predictive clustering trees with a Bayesian inference framework using variational Bayes.
Unlike traditional ensemble methods or single deterministic trees, \glspl{vspyct} explicitly model uncertainty in split parameters, providing inherent uncertainty quantification without sacrificing interpretability.
This integration represents a significant novelty by directly embedding probabilistic modeling into oblique splits, ensuring a single interpretable model can achieve performance comparable to or surpassing state-of-the-art ensembles across various prediction tasks.}

%\pbnote{Check Bayesian Additive Regression Trees (BART). Maybe add a comment/reference what is different}

@@

\added[id=r1, comment=C8]{The complete implementation is available at \url{https://github.com/viksa98/variational_spyct}.}

@@

\added[id=r2, comment=C14]{Advanced modeling methods, including fractional calculus and artificial neural networks, have been applied extensively across diverse domains for solving complex modeling and prediction problems~\cite{Abbas_2024,Abbas_2025,Ramzan_2025,Abbas_2025_2, Andonovikj_2022}.}

@@

\added[id=r1, comment=C3]{To provide a clearer overview of the related predictive clustering approaches discussed, we summarise and contrast their interpretability, capability to handle high-dimensional data, and uncertainty quantification properties in Table~\ref{tab:model_comparison}.}

\begin{table}[htbp]
\centering
\caption{Comparative summary of predictive clustering methods}
\label{tab:model_comparison}
\resizebox{\textwidth}{!}{
\begin{tabular}{lccc}
\toprule
\textbf{Model} & \textbf{Interpretability} & \textbf{Handles High-dimensional Data} & \textbf{Uncertainty Quantification}\\
\midrule
PCT~\citep{Kocev_2013} & High & Limited & No \\
OPCT~\citep{Stepisnik_2020} & Moderate to High & Moderate & No \\
SPYCT~\citep{Stepi_nik_2021} & Moderate & High & No \\
Ensemble methods & Low & High & Partial (via variance) \\
\textbf{VSPYCT (ours)} & \textbf{Moderate to High} & \textbf{High} & \textbf{Yes (inherent, Bayesian)} \\
\bottomrule
\end{tabular}
}
\end{table}

@@

\added[id=r2, comment=C8C9]{We selected Gaussian priors for split parameters due to their computational tractability and effectiveness in Bayesian inference frameworks.
The impurity function used for evaluating splits leverages the variance reduction weighted by selection probabilities, a common choice ensuring effective partitioning of data into homogeneous subsets.
Feature importance is calculated using the normalised expected weight-to-variance ratio, reflecting both the strength and reliability of each feature's contribution.}

@@

    \Procedure{LearnSplit}{$\mathcal{D}, \theta, E, \lambda, \beta, \sigma$}
        \State Initialise $\Theta = \{\boldsymbol{\mu}_w, \boldsymbol{\Sigma}_w, \mu_b, \sigma_b^2\}$ \Comment{Initialise variational parameters}
        \State Initialise optimiser: $\alpha \gets \lambda$
        \For{$e \in \{1, \dots, E\}$} \Comment{Iterate over epochs}
            \For{each mini-batch $\mathcal{B} \subseteq \mathcal{D}$ of size $\beta$}
                \State Sample $\mathbf{w} \sim \mathcal{N}(\boldsymbol{\mu}_w, \boldsymbol{\Sigma}_w)$, $b \sim \mathcal{N}(\mu_b, \sigma_b^2)$
                \State Compute impurity $\Omega(\mathbf{X}, \mathbf{Y}; \mathbf{w}, b)$ as:
                \[
                \Omega(\mathbf{X}, \mathbf{Y}; \mathbf{w}, b) = \sum_{\text{right}} \rho(\mathbf{w}^\top \mathbf{x} + b) \cdot \text{Var}_\rho(\mathbf{Y}) + \sum_{\text{left}} (1 - \rho(\mathbf{w}^\top \mathbf{x} + b)) \cdot \text{Var}_\lambda(\mathbf{Y})
                \]
                \State Set the observed impurity value: $\Omega_\text{obs} = \frac{\Omega(\mathbf{X}, \mathbf{Y}; \mathbf{w}, b)}{2}$ \Comment{Scaled to remain lower than current impurity and guide minimisation}

                \State Condition on the observed impurity using:
                \[
                \log p(\Omega_\text{obs}|\Omega(\mathbf{X}, \mathbf{Y}; \mathbf{w}, b))
                \]
                \State Compute ELBO $\mathcal{L}(\mathcal{B}; \Theta)$ as per~\eqref{eq:elbo}
                \State Compute gradient $\nabla_{\Theta} \text{ELBO}$
                \State Update $\Theta \gets \Theta + \alpha \nabla_{\Theta} \text{ELBO}$ \Comment{Gradient descent update}
            \EndFor
            \If{early stopping criterion satisfied}
                \State \textbf{break}
            \EndIf
        \EndFor
        \State \textbf{return} $\Theta$
    \EndProcedure
\end{algorithmic}
\end{algorithm}

@@

\subsection{Parameter Configuration Summary}
\added[id=r2, comment=C12]{}

Table~\ref{tab:parameter_ranges} summarises the parameter ranges tested in this study and their relevance to the optimisation of \gls{vspyct}.

\begin{table}[htbp]
\centering
\begin{tabular}{lll}
Parameter & Range tested & Description \\ \hline
Monte Carlo samples ($M$) & 10 - 100 & Approximation of posterior distributions \\
Variational iterations ($I_{vb}$) & 50 - 500 & Convergence of ELBO optimization \\
Batch size ($\beta$) & 32 - 256 & Mini-batch size for optimization \\
Learning rate ($\lambda$) & 0.001 - 0.01 & Step size for gradient updates \\
\end{tabular}
\caption{Investigated parameter configurations for VSPYCT.}
\label{tab:parameter_ranges}
\end{table}

After evaluating performance and computational trade-offs, the final chosen parameter values were set to $M=100$, $I_{vb}=100$, $\beta=128$, and $\lambda=0.01$.

@@

\added[id=r1, comment=C7]{While the present work marginalises only over split-parameter posteriors, future work will adopt Bayesian non-parametric priors over tree topologies to capture structural, as well as parameter, uncertainty. This would allow the model to reason not only about decision-boundary uncertainty, but also about the tree structure itself (i.e., how many splits to create and where to place them). Possible approaches include integrating structural priors, or developing ensemble-like variational methods that maintain a distribution over multiple plausible tree structures. Such extensions would provide a more comprehensive Bayesian treatment of decision trees and further enhance both predictive reliability and interpretability.}

\added[id=r1, comment=C6]{An explicit empirical comparison of training time and scalability between \gls{vspyct} and deterministic approaches was beyond the scope of this study, due to implementation-specific and hardware-dependent factors.}

\textit{Advantages of VSPYCT:}
\added[id=r2, comment=C10]{}
\begin{itemize}
    \item Integrated interpretability and uncertainty quantification.
    \item Competitive predictive performance compared to ensembles.
    \item Effective handling of complex classification boundaries.
\end{itemize}

\textit{Disadvantages and Limitations:}
\begin{itemize}
    \item Higher computational complexity due to variational inference.
    \item Potential scalability issues with extremely large or high-dimensional datasets.
    \item Limited capability to represent structural uncertainty compared to ensembles.
\end{itemize}

@@

\subsection{Practical Interpretation of Results}
\added[id=r2, comment=C7]{}
The empirical results presented earlier can be further understood by considering their practical implications and physical meaning in real-world contexts.
For regression tasks, such as predicting concrete compressive strength, the improved accuracy of \gls{vspyct} indicates that probabilistic modeling of uncertainty in feature weights enables better handling of measurement variability or experimental errors common in physical systems.
This modeling approach allows practitioners, for example civil engineers, to better anticipate the reliability of material properties predicted by the model.

Similarly, classification results can be interpreted from a practical perspective by analysing the decision boundaries constructed by \gls{vspyct}.
For instance, in medical diagnosis (such as chronic kidney disease), the high predictive performance reflects the model’s capacity to clearly delineate clinically meaningful groups, effectively capturing critical diagnostic thresholds that physicians rely on for clinical decision-making.

Moreover, the results obtained for employment duration prediction offer a direct practical interpretation: demographic and occupational splits made by \gls{vspyct} align closely with known socio-economic factors affecting employment outcomes.
Thus, the proposed model effectively incorporates and reveals practically meaningful dynamics from the data, enhancing both interpretability and applicability.

@@

\paragraph{Practical use for domain experts}

\added[id=r1, comment=C9]{}
After training, VSPYCT provides three complementary outputs that can be inspected with standard analysis tools:
\begin{enumerate}
  \item \textbf{Node summaries:} for every split we report the posterior mean, standard deviation, and $95\%$ credible interval of each coefficient, allowing analysts to verify which variables dominate within a specific branch.
  \item \textbf{Tree file:} the complete tree is exported in a plain, machine-readable graph format so that users can open it in any network-visualisation environment and interactively traverse, fold, or unfold decision paths.
  \item \textbf{Global relevance ranking:} the importance scores (see Figure~\ref{fig:feature_importance}) order the features by their expected effect on the prediction, helping experts prioritise factors for monitoring or intervention.
\end{enumerate}

@@

\begin{comment}
We introduced the \gls{vspyct} model, an approach that combines the benefits of oblique splits, variational Bayes, and Bayesian inference within a single, interpretable predictive framework.
The \gls{vspyct} model addresses the trade-off between the performance of ensemble models and the interpretability of single decision trees, while also tackling the absence of inherent uncertainty quantification in predictions.

@@

The model opens new possibilities for research in semi-supervised learning and Bayesian inference, promising to bridge the gap between theoretical advancements and practical applications.
Future work will focus on further refining the model's performance, exploring the inclusion of soft splits to enable smoother decision boundaries and probabilistic routing, and extending its applications in diverse predictive tasks.
\end{comment}

\added[id=r2, comment=C2C13]{}
We introduced \glspl{vspyct}, a novel framework combining Bayesian inference with predictive clustering trees.
\Glspl{vspyct} uniquely incorporate variational Bayes into oblique splits, providing built-in uncertainty quantification and preserving interpretability.

Our experimental evaluations demonstrated that \glspl{vspyct} outperform ensemble methods in binary and multi-class classification tasks, while remaining competitive in multi-target and single-target regression scenarios.
For example, \gls{vspyct} achieved the highest average rank in binary classification tasks and tied for best in multi-class classification, confirming its robustness and generalisation capabilities.

The primary advantages of \gls{vspyct} include inherent interpretability, integrated uncertainty quantification, and competitive predictive performance without requiring ensembles.
However, its main disadvantage is increased computational complexity due to variational inference optimisation and Monte Carlo sampling.

Future research should explore methods to reduce computational complexity, integrate structural priors to handle structural uncertainty, and extend \gls{vspyct} to broader domains and larger datasets, improving scalability and general applicability.

\section*{Acknowledgement}
The authors acknowledge the research core funding No.\ P2-0001, and V5-24020 financially supported by the Slovenian Research and Innovation Agency, and the Ministry of the Economy, Tourism and Sport of the Republic of Slovenia.
